---
title: "Temporal Variability of the Mexico Government COVID-19 dataset"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    toc_float: yes
    code_folding: show
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
author: |
        | Lexin Zhou^1^,  Nekane Romero^1^, Juan Martínez-Miranda^3^, J Alberto Conejero^2^, Juan M García-Gómez^1^, Carlos Sáez^1^
        | 
        | ^1^Biomedical Data Science Lab, Instituto Universitario de Tecnologías de la Información y Comunicaciones (ITACA), Universitat Politècnica de València (UPV), Camino de Vera s/n, Valencia 46022, España. ^2^Instituto Universitario de Matemática Pura y Aplicada (IUMPA), Universitat Politécnica de València, Valencia, Spain. ^3^CONACyT - Centro de Investigación Científica y de Educación Superior de Ensenada - CICESE-UT3, Mexico.
        | Corresponding authors: Lexin Zhou <lexinzhouds@gmail.com> and Carlos Sáez <carsaesi@upv.es>
date: "February 14, 2021"
---

# Introduction

In this work we describe the temporal variability analysis of the [Mexico Government dataset](https://www.gob.mx/salud/documentos/datos-abiertos-152127) as part of initial data quality assessment of the [COVID-19 Metaclustering](https://github.com/bdslab-upv/covid19-metaclustering) project of the Biomedical Data Science Lab of the Universitat Politècnica de València, Spain.

An initial assessment has shown unstable periods at the beginning and end of the repository acquisition period. This is usually related to transient states due to e.g. few cases. In this report, to improve the visualization of results we have focused on the period between 2020-03-15 and 2020-10-15.

This study can also be fully explored in our [ehrtemporalvariability.upv.es](http://ehrtemporalvariability.upv.es/) tool. Further information for EHRtemporalVariability package can be found in our GitHub repository [EHRtemporalVariability](https://github.com/hms-dbmi/EHRtemporalVariability).  .

If you use this code please cite:

<blockquote style='font-size:14px'>Lexin Zhou, Nekane Romero, Juan Martínez-Miranda, J Alberto Conejero, Juan M García-Gómez, Carlos Sáez. Heterogeneity in COVID-19 severity patterns among age-gender groups: an analysis of 778 692 Mexican patients through a meta-clustering technique. Preprint submitted to medRxiv. </blockquote>

If you are interested in collaborating in this work please [contact us](mailto:carsaesi@upv.es).

# Setup
Install and load the required packages.
```{r packages, eval=TRUE, message=FALSE, warning=FALSE}
# install.packages('EHRtemporalVariability')
# install.packages('glue')
library(EHRtemporalVariability)
library(glue)
```

# Data loading
```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
### LOAD AND FORMAT DATASET 
untar('./data/COVID19_MEXICO_LATEST_DATA.tar.gz', exdir = "data")
data = read.csv2("./data/COVID19_MEXICO_LATEST_DATA.csv", encoding="UTF-8", sep = ",", header = TRUE, na.strings = "", stringsAsFactors = FALSE,dec=".",
                 colClasses = c( "numeric",   #Index of row
                                 "Date",      #LAST_UPDATE
                                 "character", #ID_REGISTRATION
                                 "factor",    #ORIGIN
                                 "factor",    #SECTOR
                                 "character", #ENTITY_UM
                                 "factor",    #SEX
                                 "character", #ENTITY_NAT
                                 "character", #ENTITY_RES
                                 "character", #MUNICIPALITY_RES
                                 "factor",    #PATIENT_TYPE
                                 "Date",      #ADMISSION_DATE
                                 "Date",      #SYMPTOMS_DATE
                                 "Date",      #DEATH_DATE
                                 "factor",   #INTUBATED
                                 "factor",   #PNEUMONIA
                                 "integer",   #AGE
                                 "character", #NATIONALITY
                                 "factor",   #PREGNANT
                                 "factor",   #SPEAK_INDIGENOUS_LANGUAGE
                                 "factor",   #Indigenous
                                 "factor",   #DIABETES
                                 "factor",   #COPD
                                 "factor",   #ASTHMA
                                 "factor",   #INMUSUPR
                                 "factor",   #HYPERTENSION
                                 "factor",   #OTHER_DISEASE
                                 "factor",   #CARDIOVASCULAR
                                 "factor",   #OBESITY 
                                 "factor",   #CHRONIC_KIDNEY_DISEASE
                                 "factor",   #SMOKE
                                 "factor",    #OTHER_CASE_CONTACT
                                 "factor",    #TESTED
                                 "factor",    #RESULT_LAB
                                 "factor",    #FINAL_CLASIFICATION
                                 "factor",    #MIGRANT
                                 "character", #COUNTRY_NATIONALITY
                                 "character", #COUNTRY_ORIGIN
                                 "factor",    #UCI
                                 "factor",    #OUTCOME ('See the Python notebook')
                                 "numeric",   #Survival_days ('See the Python notebook')
                                 "numeric"  #From_symptoms_to_hospital_days ('See the Python notebook')
                 ))  

```

# Data preparation

```{r preprocessing, eval=TRUE, message=FALSE, warning=FALSE}
#########################################
#    SELECT ONE OF THE FOLLOWING LINE   #
#########################################
data = data[data$RESULT_LAB == 'Positive SARS-CoV-2', ]
# data = data[data$RESULT_LAB == 'Non-Positive SARS-CoV-2', ] 

# Select the variables that we want to study.
data <- select(data, LAST_UPDATE, ID_REGISTRATION, ORIGIN, SECTOR, ENTITY_UM,ENTITY_NAT,ENTITY_RES,MUNICIPALITY_RES, NATIONALITY, MIGRANT,COUNTRY_NATIONALITY, COUNTRY_ORIGIN,SEX, AGE, OBESITY, SMOKE, PREGNANT, SPEAK_INDIGENOUS_LANGUAGE, PATIENT_TYPE, ADMISSION_DATE, SYMPTOMS_DATE, DEATH_DATE,PNEUMONIA, DIABETES, COPD, ASTHMA, INMUSUPR, HYPERTENSION, CARDIOVASCULAR, CHRONIC_KIDNEY_DISEASE, OTHER_DISEASE, UCI, INTUBATED, OUTCOME, Survival_days, FromSymptomToHospital_days,OTHER_CASE_CONTACT)

# convert some variable types
data$ageCat = vector(mode = "character", length = nrow(data))
names(data)[names(data) == "UCI"] <- "ICU" # Translate UCI (Spanish) to ICU (Intensive Care Unit)
data$AGE = as.integer(data$AGE)
data$Survival_days = as.integer(data$Survival_days)
data$FromSymptomToHospital_days = as.integer(data$FromSymptomToHospital_days)

# Manually fix variable types
data$ADMISSION_DATE = as.Date(data$ADMISSION_DATE)
data$SYMPTOMS_DATE = as.Date(data$SYMPTOMS_DATE)
data$DEATH_DATE = as.Date(data$DEATH_DATE)


# remove rows with missing Admission dates and remove index column
data = data[!is.na(data$ADMISSION_DATE),-1]

# remove rows: LAST_UPDATE and ID_REGISTRATION due to their irrelevance

names(data)[names(data) == "UCI"] <- "ICU" # Translate UCI (Spanish) to ICU (Intensive Care Unit)
# remove rows with missing dates and remove patient_id column
data = subset(data, select = -c(ID_REGISTRATION))
```

# Temporal variability estimation
Estimation of the data temporal maps and information geometric temporal plot objects. We set admission date as the reference date, select a weekly analysis, and focus on the period between 2020-03-15 and 2020-10-15. To expand the period to the full timespan, remove the `startDate` and `endDate` parameters.
```{r EHRTempVar, eval=TRUE, message=FALSE, warning=FALSE}
probMaps <- estimateDataTemporalMap( data             = data, 
                                     dateColumnName   = "ADMISSION_DATE", 
                                     period           = "week",
                                     numericSmoothing = FALSE,
                                     startDate        = as.Date("2020-03-15"),
                                     endDate          = as.Date("2020-10-15"))
igtProjs <- sapply (probMaps, estimateIGTProjection, dimensions = 3)

```

# Results
```{r visualization, eval=TRUE, message=FALSE, warning=FALSE}
plotDataTemporalMap(probMaps$AGE, absolute = FALSE)
plotDataTemporalMap(probMaps$AGE, absolute = TRUE)
plotIGTProjection(igtProjs$AGE, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$SEX, absolute = FALSE)
plotIGTProjection(igtProjs$SEX, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$PATIENT_TYPE, absolute = FALSE)
plotIGTProjection(igtProjs$PATIENT_TYPE, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)

plotDataTemporalMap(probMaps$SECTOR, absolute = FALSE)
plotIGTProjection(igtProjs$SECTOR, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$ORIGIN, absolute = FALSE)
plotIGTProjection(igtProjs$ORIGIN, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$ENTITY_UM, absolute = FALSE)
plotIGTProjection(igtProjs$ENTITY_UM, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$ENTITY_RES, absolute = FALSE)
plotIGTProjection(igtProjs$ENTITY_RES, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$MUNICIPALITY_RES, absolute = FALSE, endValue = 30)
plotIGTProjection(igtProjs$MUNICIPALITY_RES, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$COUNTRY_NATIONALITY, absolute = FALSE)
plotIGTProjection(igtProjs$COUNTRY_NATIONALITY, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$MIGRANT, absolute = FALSE)
plotIGTProjection(igtProjs$MIGRANT, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)

plotDataTemporalMap(probMaps$OBESITY, absolute = FALSE)
plotIGTProjection(igtProjs$OBESITY, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$SMOKE, absolute = FALSE)
plotIGTProjection(igtProjs$SMOKE, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$PREGNANT, absolute = FALSE)
plotIGTProjection(igtProjs$PREGNANT, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)

plotDataTemporalMap(probMaps$SPEAK_INDIGENOUS_LANGUAGE, absolute = FALSE)
plotIGTProjection(igtProjs$SPEAK_INDIGENOUS_LANGUAGE, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$PNEUMONIA, absolute = FALSE)
plotIGTProjection(igtProjs$PNEUMONIA, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$DIABETES, absolute = FALSE)
plotIGTProjection(igtProjs$DIABETES, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$COPD, absolute = FALSE)
plotIGTProjection(igtProjs$COPD, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$ASTHMA, absolute = FALSE)
plotIGTProjection(igtProjs$ASTHMA, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$INMUSUPR, absolute = FALSE)
plotIGTProjection(igtProjs$INMUSUPR, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$HYPERTENSION, absolute = FALSE)
plotIGTProjection(igtProjs$HYPERTENSION, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$OTHER_DISEASE, absolute = FALSE)
plotIGTProjection(igtProjs$OTHER_DISEASE, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$CARDIOVASCULAR, absolute = FALSE)
plotIGTProjection(igtProjs$CARDIOVASCULAR, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$CHRONIC_KIDNEY_DISEASE, absolute = FALSE)
plotIGTProjection(igtProjs$CHRONIC_KIDNEY_DISEASE, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)

plotDataTemporalMap(probMaps$ICU, absolute = FALSE)
plotIGTProjection(igtProjs$ICU, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$INTUBATED, absolute = FALSE)
plotIGTProjection(igtProjs$INTUBATED, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$OTHER_CASE_CONTACT, absolute = FALSE)
plotIGTProjection(igtProjs$OTHER_CASE_CONTACT, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$OUTCOME, absolute = FALSE)
plotIGTProjection(igtProjs$OUTCOME, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$SYMPTOMS_DATE, absolute = FALSE)
plotIGTProjection(igtProjs$SYMPTOMS_DATE, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$DEATH_DATE, absolute = FALSE)
plotIGTProjection(igtProjs$DEATH_DATE, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$Survival_days, absolute = FALSE)
plotIGTProjection(igtProjs$Survival_days, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
plotDataTemporalMap(probMaps$FromSymptomToHospital_days, absolute = FALSE)
plotIGTProjection(igtProjs$FromSymptomToHospital_days, dimensions = 2, colorPalette = "Magma", trajectory = TRUE)
```

# Export for the Shiny app
You can fully explore these results in our [EHRtemporalVariability Shiny App](http://ehrtemporalvariability.upv.es/) by loading the .RData file.
```{r}
save(probMaps, igtProjs, file="variability_MexCov_Positive_weekly.RData")
```

